For Finitary Induction-Induction, \\ Induction is Enough

### 2020-05-03 paper/paper.tex
fixes

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and the main example is still the Chapman-style
\cite{chapman09eatitself} syntax for a type theory. A minimised version
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\GG:\Con)\ra\Ty\,\GG\ra\Con \\
  & \mathsf{U} && : (\GG:\Con)\ra\Ty\,\GG \\
  & \mathsf{El} && : (\GG:\Con)\ra\Ty\,(\mathsf{ext}\,\GG\,(\mathsf{U}\,\GG))
%  & \mathsf{Pi} && : (\GG:\Con)(A:\Ty\,\GG)\ra\Ty\,(\mathsf{ext}\,\GG\,A)\ra\,\Ty\,\GG
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also holds for higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

@@

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} showed a reduction in
extensional type theory however this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\GG\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle to define
$\ll\blank\rr$ is that $\ll\blank\rr$ for types refers to
$\ll\blank\rr$ for contexts, hence this function is
recursive-recursive.

Kaposi, Kov{\'a}cs, and Altenkirch
\cite{Kaposi:2019:CQI:3302515.3290315} introduced a small type theory
called the theory of signatures to describe quotient
inductive-inductive types (QIIT). QIITs are generalisations of IITs
where equality constructors are also allowed. A QIIT signature is a
context in the theory of QIIT signatures, for example natural numbers
are specified by the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of
length three ($Nat$, $zero$ and $suc$ are variable names). The theory
of QIIT signatures is itself a QIIT. In ibid., it is proved that if a
model of extensional type theory supports the theory of QIIT
signatures, then it supports all QIITs.

@@

In this paper we show that any model of extensional type theory with
indexed W-types supports the theory of IIT signatures, and as a
consequence all IITs. The difficulty in this construction is that the
theory of IIT signatures is itself a QIIT, it is both
inductive-inductive and has equality constructors. However it can be
seen as the well-typed syntax of a small type theory without any
computation rules. Hence we can represent the syntax of normal forms
without quotienting. We construct this well-typed normal syntax using
preterms and typing relations from indexed W-types. Finally, we prove
the elimination principle in the style of the initiality proof of
Streicher \cite{streichersemantics}.

@@

We describe related work in \Cref{sec:related}, and explain our
notation and Agda formalisaton in \Cref{sec:notation}. Then the
following three sections describe our three contributions:
\begin{itemize}
\item \Cref{sec:theory_of_signatures}. We define what it means for a
  model of extensional type theory (ETT, \Cref{def:ett}) to support
  all inductive-inductive types (IITs): \Cref{def:hasiits}. The novel
  contribution here is a (predicative) Church encoding of signatures
  following \cite{DBLP:conf/lics/AwodeyFS18}.
\item \Cref{sec:constructingiits}. We show that if a model of ETT
  supports the theory of IIT signatures
  (\Cref{def:theoryofsignatures}), then it supports IITs in the above
  sense: \Cref{thm:ToSToIITs}. This is an adaptation of a proof in
  \cite{Kaposi:2019:CQI:3302515.3290315}.
\item \Cref{sec:ambroise}. Our main contribution is showing that if a
  model of ETT supports indexed W-types, then it supports the theory
  of IIT signatures (\Cref{thm:WToToS}), and hence, all IITs (\Cref{thm:WToIITs}).
\end{itemize}
We list further work in \Cref{sec:conclusions}.

@@

The current work builds heavily on the work of Kaposi et al.\
\cite{Kaposi:2019:CQI:3302515.3290315} on finitary quotient
inductive-inductive types (QIITs); we reuse both QIIT syntax and
semantics by restricting to IITs, and we reuse the term model
construction of QIITs as well. We also make use of the extesion to
infinitary QIITs \cite{large_inf_qiit} to derive the specification of
the elimination principle for the theory of IIT signatures.

IITs (however not by this name) were first used to describe the
well-typed syntax of type theory \cite{nisse,chapman09eatitself}. Agda
supported these general inductive definitions even before they were
named IITs and given semantics by Nordvall Forsberg and Setzer
\cite{nordvallforsbergSetzer2010inductiveinductive}. Nordvall
Forsberg's thesis \cite{forsberg-phd} contains a specification similar
in style to Dybjer and Setzer's codes for inductive-recursive types
\cite{Dybjer99afinite}. He also develops a categorical semantics based
on dialgebras and provides a reduction of IITs to indexed inductive
types, however only constructs the simple elimination principle as
opposed to the general one. Altenkirch et al.\ \cite{gabe} define
signatures for QIITs (thus IITs as well) and their categorical
semantics, however without proving existence of initial
algebras. Their notion of signature, just as Nordvall Forsberg's
involves more encoding overhead than ours.

Cartmell \cite{gat} introduced generalised algebraic theories using a
type-theoretic syntax. Removing equations from his signatures, we
obtain finitary IIT signatures similar to ours. He does not consider
constructing initial algebras using simpler classes of inductive
types.

Hugunin \cite{jasper} constructs several IITs in cubical Agda from
inductive types. In this setting, the lack of UIP makes constructions
significantly more involved, and essentially involves
coinductive-coinductive well-formedness predicates defined as homotopy
limits. Hugunin does not consider a generic syntax of IITs and only
works on specific examples (although the examples vary greatly). He
also only constructs the simple elimination principle.

@@

Voevodsky was interested in constructing initial models of type
theories from presyntaxes. Inspired by this, Brunerie et al.\ \cite{brunerie}
formalised Streicher's proof in Agda for a type theory with $\Pi$,
$\Sigma$, $\mathbb{N}$, identity types and an infinite hierarchy of
universes. He used UIP, function extensionality and quotient types in
his formalisation. In this paper we construct a type theory without
computation rules, hence we avoid using quotients.

Intrinsic (well-typed) syntax for type theories were constructed using
IITs \cite{chapman09eatitself}, inductive-recursive types
\cite{nisse,Altenkirch:2014:CO:2631172.2631176} and QIITs
\cite{ttintt}. In this paper we avoid using such general classes of
inductive types as our goal is to reduce IITs to indexed inductive
types.

@@

\begin{definition}[Model of extensional type theory (ETT)]\label{def:ett}
  By a model of ETT we mean a category with families (CwF)
  \cite{Dybjer96internaltype,Hofmann97syntaxand} with an infinite
  predicative hierarchy of universes closed under the following type
  formers: $\Pi$, $\Sigma$, $\top$ and an identity type with
  uniqueness of identity proofs and equality reflection.
\end{definition}

@@

In this section we say what it means that a model of ETT supports
IITs. We first define the notion of IIT signature. Signatures for
algebraic theories are usually given by inductive definitions. On the
one hand, we take this even further: our notion of signature is given
by a small type theory tailer-made to describe signatures, we call
this the \emph{theory of IIT signatures}. On the other hand we would
like to avoid using a complicated inductive definition (a type theory
is a quotient inductive-inductive type \cite{ttintt}) to describe a
simpler class of inductive types. Hence we use a Church encoding
\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures thus
avoiding the need for pre-existing inductive definitions. Another
feature of our signatures is that they can include types from the
model of ETT (such as $\N$ in the
$\mathsf{isEven}$--$\mathsf{isOdd}$). This is why signatures are
specified internally to the particular model of ETT.\footnote{There is
  another method inspired by Capriotti \cite{paolo} which allows
  stating what it means that any CwF $\mathcal{C}$ (not necessarily a
  model of ETT) supports IITs with definitional computation rules. In
  this method, signatures are described in the internal language of
  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We don't
  use this approach because it is more technical, and it would not
  strengthen our main result \Cref{thm:WToIITs} as the proof of
  \Cref{thm:WToToS} needs $\mathcal{C}$ to be a model of ETT.}

@@

\begin{example}[Example contexts in a signature algebra]\label{ex:contexts}
  Given a signature algebra, we can define a context in which
  specifies natural numbers. For readability, the an informal version
  of the same context is displayed on the right using variable names.
  \[
    \boldsymbol{\cdot}\,\ext\,\U\,\ext\,z : \El\,0\,\ext\,s : 1 \Ra \El\,1
    \hspace{4.9em}
    \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
  \]
  We start with the empty context $\boldsymbol{\cdot}$, then we
  declare a sort $\U$, then we declare an operator producing an
  element of the sort denoted by $\El\,0$ where $0$ is the de Bruijn
  index referring to the sort. Finally, we declare an operator which
  takes as input an element of the sort (now it became de Bruijn index
  $1$) and produces an element of the same sort. Note the assymetry of
  the function type $\Ra$: on the left there needs to be an element of
  $\U$, while on the right any type can appear (including another
  function type). This ensures strict positivity of the operators.

  Lists with elements of a given $T : \Set$ type are given by the
  following context. Here we use the function space with external
  domain $\Ram$ to include a $T$ in the signature. For readability, we
  omit the $\lambda$ and the superscripts and we don't write the
  compatibility condition. On the right we list the same signature
  with variable names.
  \begin{alignat*}{10}
    & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
  \end{alignat*}

@@

The above examples are contexts in any signature algebra, and we could
take this as a definition of signature: $(M:\SignAlg)\ra\Con^M$ is the
usual Church-encoding of contexts. However (as we will see in
\Cref{remark:sigc}) the notion of constructor for such signatures
would be too strong. Another approach would be to assume that there is
a syntax for signature algebras (an initial signature algebra), and
then a signature would be a context in this signature algebra. We will
define syntactic signatures using this approach in the next section
(\Cref{def:syntacticsignature}), but for now we don't want to assume
the existence of any inductive type. Instead, we will use a restricted
Church encoding which needs the notion of morphism of signatures.

@@

We define IIT signatures using the Church encoding introduced by
Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
difference is that we avoid the usage of impredicativity because it is
not a problem for our use case that signatures are not in the lowest
universe.
\begin{definition}[IIT signature]\label{defn:sign}
  An IIT signature is a context in an arbitrary signature algebra that
  is compatible with morphisms:
  \begin{alignat*}{6}
    & \Sign :=\,\, && \big({sig}:(M:\SignAlg)\ra\Con^M\big)\times \\
    & && \big((M\,N:\SignAlg)(f:\SignMor\,M\,N)\ra f_\Con\,({sig}\,M) = {sig}\,N \big).
  \end{alignat*}
\end{definition}
The compatibility condition says that if we obtain an $M$-context
using ${sig}$ in model $M$ and then we transport it to $N$ using $f$,
we get the same $N$-context as directly applying ${sig}$ to $N$.

Our notion of signatures don't form a signature algebra.
\begin{lemma}
  There is no $M : \SignAlg$, in which $\Con^M = \Sign$.
\end{lemma}
\begin{proof}
  If the $\Con$ component in $\SignAlg$ is $\Set_i$, then $\SignAlg$
  is in $\Set_{i+1}$, but as $\Sign$ is defined as
  $(\SignAlg\ra\dots)\times\dots$, it is at least in $\Set_{i+1}$, so
  we can't choose $\Con^M : \Set_i$ to be $\Sign : \Set_{i+1}$.
\end{proof}

Note that the notion of IIT signature is relative to a model of ETT:
it is expressed as a term (of a function type) in the model. This is
necessary because of the function space $\Pim$ which has as domain an
arbitrary type in the model. We make use of $\Pim$ in signatures with
external parameters, like the type of the elements in lists.

@@

We first explain it by columns: the first column ($^\A$ components)
corresponds to the standard model (set model, metacircular
interpretation \cite{ttintt}): contexts are sets, types are families,
terms are functions, the universe $\U$ is given by $\Set$, function
spaces are given by the external function space. The $^\D$ column is a
logical predicate interpretation, $^\A$ and $^\D$ together are a unary
version of the parametric model for dependent types
\cite{10.1145/2535838.2535852}. Contexts are predicates, types are
families of predicates, terms say that the $^\A$ interpretation
respects the predicates (this is ususally called fundamental lemma of
the logical predicate). $\U$ is given by predicate space, the
predicate at a $\Pi$ type holds for a function if it respects the
predicates. For $\Pim$, the predicate is defined pointwise. The last
column $^\S$ is a modified dependent logical relation which refers to
both $^A$ and $^\D$. Contexts are binary relations where the second
parameter depends on the first one, types are dependent variants of
this, terms say that the relation is respected by $^\A$ and $^\D$,
respectively. $\U$ is however not relation space, but a function and
$(\El\,a)^\S$ is the graph of the function $a^\S$. $\Pi^\S$ for a
function again says that the function respects the relation, however
we don't simply say
\[
  (\Pi\,a\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:a^\A\,\gamma)(\alpha^D:a^\D\,\gamma^D\,\alpha)(\alpha^\S:(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D)\ra B^\S\,\dots,
\]
as $(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D$ is just an equality
$a^\S\,\gamma^S\,\alpha = \alpha^D$ which we can singleton
contract. So we omit $\alpha^D$ and this equality as an input and
replace $\alpha^D$ by $a^\S\,\gamma^S\,\alpha$ in the definition.

@@

In the previous section, using the notions of signature algebras and
signature morphisms, we defined IIT signatures and what it means for a
model of ETT to support all IITs. In this section we show that if a
model of ETT supports the theory of IIT signatures, then it supports
all IITs. In the sense of the Church encoding of \Cref{defn:sign},
every model of ETT can describe ITT signatures. In contrast, in
\Cref{def:theoryofsignatures}, we will require existence of an initial
signature algebra.

@@

In this section we show that any model of ETT which supports indexed
W-types also supports the theory of signatures, and as a consequence
of \Cref{thm:ToSToIITs}, all IITs. For this, we work in the internal
language of a model of ETT supporting indexed W-types
\cite{indexedcont}. Indexed W-types correspond to the usual notion of
(possibly mutual) indexed inductive types. We will use Agda-style
notation for defining such inductive families: we list the sorts and
constructors and use pattern matching when eliminating from them. For
an encoding of mutual inductive families as families as indexed
W-types, see e.g.\ \cite{mutual}.

@@

\section{Conclusions}
\label{sec:conclusions}

>>>>

Inductive-inductive types (IITs \cite{forsberg-phd}) allow the mutual
definition of a type and a family of types over the first one. IITs
were originally introduced to represent the well-typed syntax of type
theory itself, and a prominent example is still Chapman's
\cite{chapman09eatitself} syntax for a type theory. A minimised version
is the IIT of contexts and types given by the following constructors.
\begin{alignat*}{5}
  & \Con && : \Set \\
  & \Ty && : \Con\ra\Set \\
  & \mathsf{empty} && : \Con \\
  & \mathsf{ext} && : (\GG:\Con)\ra\Ty\,\GG\ra\Con \\
  & \mathsf{U} && : (\GG:\Con)\ra\Ty\,\GG \\
  & \mathsf{El} && : (\GG:\Con)\ra\Ty\,(\mathsf{ext}\,\GG\,(\mathsf{U}\,\GG))
%  & \mathsf{Pi} && : (\GG:\Con)(A:\Ty\,\GG)\ra\Ty\,(\mathsf{ext}\,\GG\,A)\ra\,\Ty\,\GG
\end{alignat*}
This type has two sorts, $\Con$ and $\Ty$. The $\mathsf{ext}$ constructor of
$\Con$ refers to $\Ty$ and the $\Ty$-constructor $\mathsf{U}$ refers to $\Con$,
hence the two sorts have to be defined simultaneously. Moreover, $\Ty$ is
indexed over $\Con$. This precludes a reduction analogous to the reduction of
$\mathsf{isEven}$--$\mathsf{isOdd}$, as we would get a type indexed over
itself. Another unique feature of IITs (which also holds for higher inductive
types \cite{HoTTbook}) is that later constructors can refer to previous
constructors: in our case, $\mathsf{El}$ mentions
$\mathsf{ext}$.
% \[
%   \mathsf{Con?} : (b:\mathsf{Bool})\ra\mathsf{if}\,b\,\mathsf{then}\,\Set\,\mathsf{else}\,(\mathsf{Con?}\,\mathsf{false}\ra\Set),
% \]

@@

Reducing IITs to inductive types (more precisely, to indexed W-types)
is an open problem. Forsberg \cite{forsberg-phd} presented a reduction in
extensional type theory, however, this only provides a simpler,
non-recursive-recursive elimination principle. Hugunin \cite{jasper}
reduced several IITs to inductive types, working inside a cubical type
theory, but he also only constructed the simple eliminator. To
illustrate the difference, we list the motives and methods and the
simple elimination principle for the $\Con$--$\Ty$ example. Again, we
use implicit quantifications.
\begin{alignat*}{5}
  & Con^S && : \Con\ra\Set \\
  & Ty^S && : \Ty\,\GG\ra\Set \\
  & {empty}^S && : Con^S\,\mathsf{empty} \\
  & {ext}^S && : Con^S\,\GG\ra Ty^S\,\,A\ra Con^S\,(\mathsf{ext}\,\GG\,A) \\
  & U^S && : Con^S\,\GG\ra Ty^S\,(\mathsf{U}\,\GG) \\
  & El^S && : Con^S\,\GG\ra\,Ty^S\,(\mathsf{El}\,\GG) \\
  & \mathsf{selimCon} && : (\GG:\Con)\ra Con^S\,\GG \\
  & \mathsf{selimTy} && : (A:\Ty\,\GG)\ra Ty^S\,A
\end{alignat*}
This simple elimination principle is too weak, e.g.\ to define the
standard (metacircular) interpretation of our small syntax
\cite{ttintt}. Using pattern matching notation, this interpretation is
the following:
\begin{alignat*}{5}
  & \ll\blank\rr && : \Con\ra\Set_1 \\
  & \ll\blank\rr && : \ll\GG\rr\ra\Set_1 \\
  & \ll\mathsf{empty}\rr && := \top \\
  & \ll\mathsf{ext}\,\GG\,A\rr && := (\gamma:\ll\GG\rr)\times\ll A\rr\,\gamma \\
  & \ll\mathsf{U}\,\GG\rr\,\gamma && := \Set \\
  & \ll\mathsf{El}\,\GG\rr\,(\gamma,X) && := X
\end{alignat*}
The reason that we need the general elimination principle to define
$\ll\blank\rr$ is that $\ll\blank\rr$ for types refers to
$\ll\blank\rr$ for contexts, hence this function is
recursive-recursive.

Kaposi, Kov{\'a}cs, and Altenkirch
\cite{Kaposi:2019:CQI:3302515.3290315} introduced a small type theory,
 called the theory of signatures, to describe quotient
inductive-inductive types (QIIT). QIITs are generalisations of IITs
where equality constructors are also allowed. A QIIT signature is a
context in the theory of QIIT signatures, for example natural numbers
are specified by the context $(Nat:\U,zero:Nat,suc:Nat\ra Nat)$ of
length three ($Nat$, $zero$ and $suc$ are variable names). The theory
of QIIT signatures is itself a QIIT. In ibid., it is proved that if a
model of extensional type theory supports the theory of QIIT
signatures, then it supports all QIITs.

@@

In this paper we show that any model of extensional type theory with
indexed W-types supports the theory of IIT signatures, and as a
consequence all IITs. The difficulty in this construction is that the
theory of IIT signatures is itself a QIIT, it is both
inductive-inductive and has equality constructors. However, it can be
seen as the well-typed syntax of a small type theory without any
computation rules. Hence we can represent the syntax of normal forms
without quotienting. We construct this well-typed normal syntax using
preterms and typing relations from indexed W-types. Finally, we prove
the elimination principle in the style of the initiality proof of
Streicher \cite{streichersemantics}.

@@

We describe related work in \Cref{sec:related}, and explain our
notation and Agda formalisaton in \Cref{sec:notation}. Then the
following three sections describe our three contributions:
\begin{itemize}
\item \Cref{sec:theory_of_signatures}. We define what it means for a
  model of extensional type theory (ETT, \Cref{def:ett}) to support
  all inductive-inductive types (IITs): \Cref{def:hasiits}. The novel
  contribution here is a (predicative) Church encoding of signatures
  following \cite{DBLP:conf/lics/AwodeyFS18}.
\item \Cref{sec:constructingiits}. In \Cref{thm:ToSToIITs}, we show that if a
  model of ETT supports the theory of IIT signatures
  (\Cref{def:theoryofsignatures}), then it supports IITs in the previously
  given. This is an adaptation of a proof in
  \cite{Kaposi:2019:CQI:3302515.3290315}.
\item \Cref{sec:ambroise}. Our main contribution is showing that if a
  model of ETT supports indexed W-types, then it supports the theory
  of IIT signatures (\Cref{thm:WToToS}), and hence, all IITs (\Cref{thm:WToIITs}).
\end{itemize}
We list further work in \Cref{sec:conclusions}.

@@

The current work builds heavily on the work of Kaposi et al.\
\cite{Kaposi:2019:CQI:3302515.3290315} on finitary quotient
inductive-inductive types (QIITs); we reuse both QIIT syntax and
semantics by restricting to IITs, and we reuse the term model
construction of QIITs as well. We also make use of the extension to
infinitary QIITs \cite{large_inf_qiit} to derive the specification of
the elimination principle for the theory of IIT signatures.

IITs (although not by this name) were first used to describe the
well-typed syntax of type theory \cite{nisse,chapman09eatitself}. Agda
supported these general inductive definitions even before they were
named IITs and given semantics by Nordvall Forsberg and Setzer
\cite{nordvallforsbergSetzer2010inductiveinductive}. Nordvall
Forsberg's thesis \cite{forsberg-phd} contains a specification similar
in style to Dybjer and Setzer's codes for inductive-recursive types
\cite{Dybjer99afinite}. He also develops a categorical semantics based
on dialgebras and provides a reduction of IITs to indexed inductive
types, however only constructs the simple elimination principle as
opposed to the general one. Altenkirch et al.\ \cite{gabe} define
signatures for QIITs (thus IITs as well) and their categorical
semantics, however without proving existence of initial
algebras. Their notion of signature, like Nordvall Forsberg's,
involves more encoding overhead than ours.

Cartmell \cite{gat} introduced generalised algebraic theories using a
type-theoretic syntax. Removing equations from his signatures and only
considering finite signatures, we obtain finitary IIT signatures similar to
ours. He does not consider constructing initial algebras using simpler classes
of inductive types.

Hugunin \cite{jasper} constructs several IITs in cubical Agda from
inductive types. In this setting, the lack of UIP makes constructions
significantly more involved, and essentially involves
coinductive-coinductive well-formedness predicates defined as homotopy
limits. Hugunin does not consider a generic syntax of IITs and only
works on specific examples (although the examples vary greatly). He
also only constructs simple elimination principles.

@@

Voevodsky was interested in constructing initial models of type
theories from presyntaxes. Inspired by this, Brunerie et al.\ \cite{brunerie}
formalised Streicher's proof in Agda for a type theory with $\Pi$,
$\Sigma$, $\mathbb{N}$, identity types and an infinite hierarchy of
universes. They used UIP, function extensionality and quotient types in
the formalisation. In this paper we construct a type theory without
computation rules, hence we avoid using quotients.

Intrinsic (well-typed) syntaxes for type theories were constructed using
IITs \cite{chapman09eatitself}, inductive-recursive types
\cite{nisse,Altenkirch:2014:CO:2631172.2631176} and QIITs
\cite{ttintt}. In this paper we avoid using such general classes of
inductive types as our goal is to reduce IITs to indexed inductive
types.

@@

\begin{definition}[Model of extensional type theory (ETT)]\label{def:ett}
  By a model of ETT we mean a category with families (CwF)
  \cite{Dybjer96internaltype,Hofmann97syntaxand} with a countable
  predicative hierarchy of universes closed under the following type
  formers: $\Pi$, $\Sigma$, $\top$ and an identity type with
  uniqueness of identity proofs and equality reflection.
\end{definition}

@@

In this section we specify what it means that a model of ETT supports
IITs. We first define the notion of IIT signature. Signatures for
algebraic theories are usually given by inductive definitions. On the
one hand, we take this even further: our notion of signature is given
by a small type theory tailor-made to describe signatures, which we call
the \emph{theory of IIT signatures}. On the other hand we would
like to avoid using a complicated inductive definition (a type theory
is a quotient inductive-inductive type \cite{ttintt}) to describe a
simpler class of inductive types. Hence we use a Church encoding
\cite{DBLP:conf/lics/AwodeyFS18} of the theory of IIT signatures, thereby
avoiding the need for pre-existing inductive definitions. Another
feature of our signatures is that they can include types from the
model of ETT (such as $\N$ in the
$\mathsf{isEven}$--$\mathsf{isOdd}$). This is why signatures are
specified internally to the particular model of ETT.\footnote{There is
  another method inspired by Capriotti \cite{paolo} which allows
  stating what it means that any CwF $\mathcal{C}$ (not necessarily a
  model of ETT) supports IITs with definitional computation rules. In
  this method, signatures are described in the internal language of
  $\hat{\mathcal{C}}$, the presheaf model over $\mathcal{C}$. We do not
  use this approach because it is more technical, and it would not
  strengthen our main result \Cref{thm:WToIITs} as the proof of
  \Cref{thm:WToToS} needs $\mathcal{C}$ to be a model of ETT.}

@@

\begin{example}[Example contexts in a signature algebra]\label{ex:contexts}
  Given a signature algebra, we can define a context which
  specifies natural numbers. For readability, an informal version
  of the same context is displayed on the right using variable names.
  \[
    \boldsymbol{\cdot}\,\ext\,\U\,\ext\,z : \El\,0\,\ext\,s : 1 \Ra \El\,1
    \hspace{4.9em}
    \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
  \]
  We start with the empty context $\boldsymbol{\cdot}$, then we
  declare a sort $\U$, then we declare an operator producing an
  element of the sort denoted by $\El\,0$ where $0$ is the de Bruijn
  index referring to the sort. Finally, we declare an operator which
  takes as input an element of the sort (now it became de Bruijn index
  $1$) and produces an element of the same sort. Note the asymmetry of
  the function type $\Ra$: the domain needs to be an element of
  $\U$, while the codomain can be any type (including another
  function type). This ensures strict positivity of the operators.

  Lists with elements of a given $T : \Set$ type are given by the
  following context. Here we use the function space with external
  domain $\Ram$ to include a $T$ in the signature. For readability, we
  omit the $\lambda$ and the superscripts and we do not write the
  compatibility condition. On the right we list the same signature
  with variable names.
  \begin{alignat*}{10}
    & \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,T\Ram 1\Ra \El\,1 \hspace{3.56em} \boldsymbol{\cdot}\,\ext\,L:\U\,\ext\,nil : \El\,L\,\ext\,cons : T\Ram L\Ra \El\,L
  \end{alignat*}

@@

The above examples are contexts in any signature algebra, and we could
take this as a definition of signature: $(M:\SignAlg)\ra\Con^M$ is the
usual Church-encoding of contexts. However (as we will see in
\Cref{remark:sigc}) the notion of constructor for such signatures
would be too strong. Another approach would be to assume that there is
a syntax for signature algebras (an initial signature algebra), and
then a signature would be a context in this signature algebra. We will
define syntactic signatures using this approach in the next section
(\Cref{def:syntacticsignature}), but for now we do not want to assume
the existence of any inductive type. Instead, we will use a restricted
Church encoding. This requires the notion of morphism of signatures.

@@

We define IIT signatures using the Church encoding introduced by Awodey, Frey
and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A difference is that we avoid
impredicativity. This restricts the possible eliminations on signatures: we can
only eliminate into a universe which is smaller than the level of
signatures. However, this still covers all eliminations in this paper, and it is
also not an issue for us that signatures do not live in the smallest universe.

%% We define IIT signatures using the Church encoding introduced by
%% Awodey, Frey and Speight \cite{DBLP:conf/lics/AwodeyFS18}. A
%% difference is that we avoid the usage of impredicativity because it is
%% not a problem for our use case that signatures are not in the lowest
%% universe.
\begin{definition}[IIT signature]\label{defn:sign}
  An IIT signature is a context in an arbitrary signature algebra, which is also
  compatible with morphisms:
  \begin{alignat*}{6}
    & \Sign :=\,\, && \big({sig}:(M:\SignAlg)\ra\Con^M\big)\times \\
    & && \big((M\,N:\SignAlg)(f:\SignMor\,M\,N)\ra f_\Con\,({sig}\,M) = {sig}\,N \big).
  \end{alignat*}
\end{definition}
The compatibility condition says that if we obtain an $M$-context
using ${sig}$ in model $M$ and then we transport it to $N$ using $f$,
we get the same $N$-context as directly applying ${sig}$ to $N$.

The lack of impredicativity implies that our notion of signatures do not form a
signature algebra.
\begin{lemma}
  There is no $M : \SignAlg$, in which $\Con^M = \Sign$.
\end{lemma}
\begin{proof}
  If the $\Con$ component in $\SignAlg$ is $\Set_i$, then $\SignAlg$
  is in $\Set_{i+1}$, but as $\Sign$ is defined as
  $(\SignAlg\ra\dots)\times\dots$, it is at least in $\Set_{i+1}$, so
  we can't choose $\Con^M : \Set_i$ to be $\Sign : \Set_{i+1}$.
\end{proof}

Note that the notion of IIT signature is relative to a model of ETT:
it is expressed as a term (of a function type) in the model. This is
necessary because of the function space $\Pim$, which has as domain an
arbitrary type in the model. We make use of $\Pim$ in signatures with
external parameters, like the type of the elements in lists.

@@

We first explain it by columns: the first column ($^\A$ components)
corresponds to the standard model (set model, metacircular
interpretation \cite{ttintt}): contexts are sets, types are families,
terms are functions, the universe $\U$ is given by $\Set$, function
spaces are given by the external function space. The $^\D$ column is a
logical predicate interpretation, $^\A$ and $^\D$ together are a unary
version of the parametric model for dependent types
\cite{10.1145/2535838.2535852}. Contexts are predicates, types are
families of predicates, terms say that the $^\A$ interpretation
respects the predicates (this is ususally called fundamental lemma of
the logical predicate). $\U$ is given by predicate space, the
predicate at a $\Pi$ type holds for a function if it respects the
predicates. For $\Pim$, the predicate is defined pointwise. The last
column $^\S$ is a modified dependent logical relation which refers to
both $^A$ and $^\D$. Contexts are binary relations where the second
parameter depends on the first one, types are dependent variants of
this, terms say that the relation is respected by $^\A$ and $^\D$,
respectively. $\U$ is however not relation space, but a function and
$(\El\,a)^\S$ is the graph of the function $a^\S$. $\Pi^\S$ for a
function again says that the function respects the relation, however
we do not simply say
\[
  (\Pi\,a\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:a^\A\,\gamma)(\alpha^D:a^\D\,\gamma^D\,\alpha)(\alpha^\S:(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D)\ra B^\S\,\dots,
\]
as $(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D$ is just an equality
$a^\S\,\gamma^S\,\alpha = \alpha^D$ which we can singleton
contract. So we omit $\alpha^D$ and this equality as an input and
replace $\alpha^D$ by $a^\S\,\gamma^S\,\alpha$ in the definition.

@@

In the previous section, using the notions of signature algebras and
signature morphisms, we defined IIT signatures and what it means for a
model of ETT to support all IITs. In this section we show that if a
model of ETT supports the theory of IIT signatures, then it supports
all IITs. Using the Church encoding of \Cref{defn:sign},
every model of ETT can describe ITT signatures. In contrast, in
\Cref{def:theoryofsignatures}, we will require existence of an initial
signature algebra.

@@

In this section we show that any model of ETT which supports indexed W-types
also supports the theory of signatures, and as a consequence of
\Cref{thm:ToSToIITs}, all IITs. For this, we work in the internal language of a
model of ETT supporting indexed W-types \cite{indexedcont}. Indexed W-types
correspond to the usual notion of (possibly mutual) indexed inductive types. We
use Agda-style notation to define such inductive families: we list the sorts and
constructors and use pattern matching when eliminating from them. For an
encoding of mutual inductive families as indexed W-types, see
e.g.\ \cite{mutual}.

@@

\section{Further Work}
\label{sec:conclusions}
